\section{Introduction}
%\subsection*{Performance models are important}
Most software systems can be customized via configuration options to meet user demands. The selection of configuration options can enable desired functionality (features) or tweak non-functional aspects of a software system, such as improving performance or energy consumption. 
All too often, software performance bugs can be linked to configuration options~\cite{han_empirical_2016}. 
The relationship of configuration choices and their influence on non-functional properties has been extensively studied in recent years. Most work focuses on the prediction of non-functional properties\footnote{In this paper, we use the terms \textit{non-functional properties} and \textit{performance} interchangably, although the latter is more restricive. In the context of this work a performance model describes any model that predicts one or more non-functional properties.}, such as execution time or memory utilization, for arbitrary configurations~\cite{dorn2020,siegmundPerformanceinfluenceModelsHighly2015,haDeepPerf2019,perfAL,guoVariabilityawarePerformancePrediction2013,sarkarCostEfficientSamplingPerformance,guo_2018_data,fourier_learning_2015,perLasso} or on finding configurations that are near-optimal with regard to a non-functional property~\cite{nairUsingBadLearners2017,nairFlash18,ohFindingNearoptimalConfigurations2017}. 
The backbone of a performance estimate is a model that maps a given configuration to the estimated performance value. 
In general, performance prediction models are only useful to practitioners if predictions hold for a large number of application scenarios.

%\subsection*{Performance models can be workload-biased}
Learning performance models for software performance usually relies on a training set of configuration-specific performance measurements. 
Typically, established approaches employ a single workload (e.g., a benchmark or test suite) to measure performance, which usually aims at emulating a real-world application scenario. 
However, as the collected measurements represent only the one application scenario, the resulting performance model can  estimate performance for this single scenario reliably. 
The choice of a single workload can introduce bias as the predictive model is only as useful as the underlying workload resembles a wide range of use cases in the real world. 
Pereira et al. illustrate this fact with an analysis of the video encoder \textsc{x264} under different workloads~\cite{alves_sampling_2020}. 
For two different workloads, they observed vastly different performance distributions across a large set of configurations.
While this particular observation can be an outlier, in general, one cannot assume that performance behavior under a single workload is generalizable, and therefore consistent in every real-world setting.
In this vein, a recent exploratory study by Jamshidi et al.\cite{jamishidi_transfer_2017} illustrates that performance models of a configurable software system often exhibit differences in estimated performance influences when learned in different environments, including different versions, hardware setups, and workloads.
Performance  influences across different environments remain largely congruent and differences only affect small numbers of configuration options, such that differences can be learned efficiently via transfer learning~\cite{jamshidi_transfer_gp_2017,jamshidi_learning_2018}. 
Nonetheless, under a single workload, we are left with performance models whose estimations are potentially inaccurate for different environments.

%\subsection*{Towards Estimating Representativeness}
Given not one, but a set of performance models based on different workloads, the question arises: Which performance model of the set is most general, whose estimations are more likely to \emph{represent} real-world scenarios? 
To assess which model's estimations to trust most, we require information about whether and how different workloads interact with the software system under test. 

Take as an example the method in Listing~\ref{lst:intro} from an imaginary database system. It receives as an argument an array of strings, each a row to insert. The method exhibits two configuration options, of whose selection some code sections depend: if \texttt{DUPLICATE\_CHECK} is enabled, the passed array is checked for duplicates via a separate method invocation (lines 2--4), and if \texttt{AUTOCOMMIT} is enabled, each insertion is treated as a single transaction (lines 10--12). The latter section, however, is not only depends on the selection of a configuration option, but also on the number of rows to insert. If the workload (here, the number of rows) is large enough, the individual insertion per row (line 9) is replaced by another insertion method (line 6). This example illustrates, how the execution of a program could depend on both configuration decisions (lines 2--4) and workloads (lines 5--9), or interactions between them (lines 10--12).

\begin{lstlisting}[caption={Illustrative code example of option-dependent code, workload-dependent code, and interaction-dependent code.},label=lst:intro, escapechar=\%,float]
public void insertRows(String[] rows) {
%\cova%	if (Configuration.DUPLICATE_CHECK) {
%\cova%		rows = this.removeDuplicates(rows);
%\cova%	}
%\covc%	if (rows.length() > 50) {
%\covc%		this.insertBulkRows(rows);
%\covc%	} else {
%\covc%		for (String row: rows) {
%\covc%			this.insertRow(row);
%\covb%			if (Configuration.AUTOCOMMIT) {
%\covb%				this.commit();
%\covb%			}	
%\covc%		}
%\covc%	}
}
\end{lstlisting}

To now representatively assess the method's performance, one would require not only observations under a set of sample configurations, but also under different workloads to cover and measure performance for workload-dependent code. If dependent code, such as the interaction between \texttt{AUTOCOMMIT} and the workload size is missed, the estimation of an option's performance influence becomes unrepresentative and, thus, the performance model unreliable.

%\subsection*{Approach, Results \& Contribution}
In this paper, we explore how the selection of a workload can influence performance models how prone these are to be distorted. To this end, we extend our perspective from pure black-box measurements~\cite{dorn2020,siegmundPerformanceinfluenceModelsHighly2015,haDeepPerf2019,perfAL,guoVariabilityawarePerformancePrediction2013,sarkarCostEfficientSamplingPerformance,guo_2018_data,fourier_learning_2015,perLasso} with knowledge about the code executed. Specifically, we enrich performance measurements from different configurations under varying workloads with corresponding easy-to-compute code coverage data. 
We conduct an empirical study of seven configurable software systems to explore how the selection of workloads can influence both performance and code coverage, and possibly distort performance models (cf. Section~\ref{sec:study}). {\color{teal}We found that \ldots.} 
Based on our findings, we explore correlations between code coverage and differences in performance influence (Section~\ref{sec:metric}). {\color{teal}	We found that \ldots. We propose to use code coverage as a \ldots and devise a novel method to interpret performance model with regard to code coverage in order to estimate their representativeness.}

In summary, we offer the following contributions:
\begin{itemize}
	\item A companion Web site\footnote{\url{https://github.com/icse-submitter123/submission}} providing supplementary material including performance measurements, code coverage data, and additional visualizations.
\end{itemize}
\section{Background and Related Work}
\subsection{Performance Models for Configurable Software}
Configurable software systems are an umbrella term for any kind of software system that exhibits configuration options to customize functionality. 
While the primary purpose of configuration options is to select (categorical or binary options) and tune (numerical options) functionality, each configuration choice may also have implications on non-functional properties---be it intentional or unintentional. Finding configurations with optimal performance~\cite{nairUsingBadLearners2017,nairFlash18,ohFindingNearoptimalConfigurations2017} and estimating the performance for arbitrary configurations is an established line of research\cite{siegmundPerformanceinfluenceModelsHighly2015,haDeepPerf2019,perfAL,guoVariabilityawarePerformancePrediction2013,sarkarCostEfficientSamplingPerformance,guo_2018_data,fourier_learning_2015,perLasso}. 

For the latter, machine-learning techniques are used to learn \emph{performance models} that approximate non-functional properties, such as execution time or memory usage, as a function of software configurations $c \in C$, formally $\Pi: C \rightarrow \mathbb{R}$.
Performance models can be obtained using a variety of machine-learning techniques, including probabilistic programming~\cite{dorn2020}, multiple linear regression~\cite{siegmundPerformanceinfluenceModelsHighly2015}, classification and regression trees~\cite{sarkarCostEfficientSamplingPerformance,guo_2018_data}, Fourier learning~\cite{fourier_learning_2015,perLasso}, and deep neural networks~\cite{haDeepPerf2019,perfAL}.
The set of configurations for training can be sampled from the configuration space using a variety of different sampling techniques. All sampling strategies aim at yielding a representative sample, either by covering main effects of configuration options and interactions among them~\cite{siegmundPredictingPerformanceAutomated2012}, or sampling uniformly from the configuration space~\cite{ohFindingNearoptimalConfigurations2017,kaltenecker_distance-based_2019}.

Most approaches share the perspective of treating a configurable software system as a black box model at application-level granularity. Besides, recent work has incorporated feature location techniques to guide sampling effort towards relevant configuration options~\cite{velez_2020_configcrusher_jase,velez_comprex_2021} or model non-functional properties similarly, but at finer granularity~\cite{weber_white_2021}.

\subsection{Performance under Varying Workloads}
Differences in the performance behavior of configurable software systems have been documented before~\cite{jamishidi_transfer_2017,alves_sampling_2020}. Jamshidi et al. observe that performance influences across different environments remain largely congruent and differences only affect small numbers of configuration options, such that differences can be learned efficiently via transfer learning~\cite{jamishidi_transfer_2017,jamshidi_learning_2018,jamshidi_transfer_gp_2017,ding_bayesian_2020}. Here, the bias is explicitly learned to adapt an already existing – and potentially also biased -- performance model. The benefit of using more than one specific workload for performance models is illustrated by work of Liao et al.~\cite{liao_2020_using_emse}. For two different versions of a software system, they learned performance models based on disjoint sets of workloads. From the comparison of both performance models, they were able to identify performance shifts as the workload bias was averaged out.
Closely related to our work, aspects of workload variability and bias have been studied for micro-benchmarks, such as test suites for performance testing. Laaber et al. use static analysis to estimate the stability (convergence after repeated measurements) of a given micro-benchmark based on usage of specific code features~\cite{laaber_emse_2021}. Similarly to our problem, yet for micro-benchmarks instead of system-level workloads, Grambow et al. use execution logs to infer redundant or missing coverage~\cite{grambow_peerj_2021}. With this approach, they are able to condense benchmark suites to select only relevant ones and also recommend code sections to test with further workloads. In contrast to our work, this does not consider configuration options.
%\pdfcomment[color=green!25]{
%Die lernen aber ein Performance-Model nicht anhand von Konfigurationen so wie wir es tun, %sondern anhand von Logs, um auch Änderungen der Workloads leicht abdecken zu können. Ich finde, dass man das verdeutlichen sollte.
%}


\subsection{Associating Code and Features}\label{sec:feature_location}
The problem of determining what code sections implement certain functionality in a software system is known as \emph{feature location}~\cite{rubin_feature_2013}. To reason about the degree to what a software feature (i.e., code conditioned by a configuration option) is covered under a workload, knowledge of corresponding code regions is necessary and can be obtained from data and control-flow analysis as well as dynamic analyses.

Most feature location approaches either employ static or dynamic program analysis to infer an association between code and features. For static analyses, variables that encode configuration options (e.g., \texttt{AUTOCOMMIT} in Listing~\ref{lst:intro}) are tainted and from following data-flow and control-flow, one can infer code sections tainted by such variables~\cite{velez_2020_configcrusher_jase,lillack_2018_lotrack_tse,luo_2019_cova}.
While static analyses can yield precise results, scalability is often limited by the exploration of possible execution paths. To mitigate this shortcoming, dynamic taint analysis taints variables similarly to the above works, but only follows one execution path~\cite{bell_phosphor_2014,velez_comprex_2021,splat_kim_2013}. From few executions of different configurations one can then extract feature-specific code from the code that was covered. In the context of the example in Listing~\ref{lst:intro}, two runs with \texttt{DUPLICATE\_CHECK} either enabled or disabled, respectively, would allow to infer dependent lines of code.


Aside from program analysis, a more tame, but also less precise approach is to use code coverage information, such as execution traces.
The rationale is that by exercising feature code, for instance via enabling configuration options or running corresponding tests, its location can be inferred from code coverage differences. Applications of such an approach have been studied not only for feature location~\cite{wong_integrated_2005,sulir_annotation_2015,michelon_spectrum_2021,perez_framing_2016}, but root in program comprehension~\cite{wilde_early_1996,wilde_reconnaissance_1995,sherwood_reducing_nodate,perez_diagnosis_2014,castro_pangolin_2019} and fault localization~\cite{agrawal_fault_1995,wong_faultloc_2016}. In the context of our example from Listing~\ref{lst:intro}, the feature code for option \texttt{DUPLICATE\_CHECK} could be inferred from the diff between execution traces where that option is once enabled and once disabled.

\section{Empirical Study}~\label{sec:study}

%\subsection{Exploring Workload-specific Differences}
Workload selection can determine the coverage of code regions, but this does not necessarily translate to observable shifts in performance and vice versa. Nonetheless, both metrics, code coverage and performance, in isolation are used to make statements about the quality of a software system. We argue that, even without drawing a connection between both, exploring the impact that the selection of different workloads can have is worthwhile, for instance to select for stable workloads or robust configurations.

In the first part of our contribution, hence, we explore the influence that different workloads can have on performance behavior, as well as code coverage. To better understand the characteristics and observations of the subject systems in question, we first ask the following two research questions:

\RQ{1.1}{How similar are spectra of configuration performance across workloads?}

\RQ{1.2}{How similar is code coverage of configurations across workloads?}

As previous work has already demonstrate workload-specific differences in performance spectra (performance distribution of a set of configurations)~\cite{alves_sampling_2020}, we first revisit this aspect and study how workload differences influence the distribution of both performance (\RQref{1.1}) and code coverage (\RQref{1.2}).

Given the presence of workload-specific differences in both code coverage and performance of a configurable software system, we subsequently explore whether such differences could be attributed to workloads as a common factor of variation. The rationale is that an existing correlation between differences in both dependent variables could allow for using code coverage as a proxy/indicator for differences in performance. To this end, we ask the following research question:

\RQ{1.3}{Do workload-specific differences in code coverage correspond to workload-specific differences in performance?}

\begin{table*}[ht!]
	\centering
	\caption{Subject System Characteristics}
	\input{tables/subject_systems.tex}
	\label{tab:subject_systems}
\end{table*}

\subsection{Experiment Setup}
\subsubsection{Subject Systems \& Workloads}

For our study, we have selected seven configurable software systems as illustrated in Table~\ref{tab:subject_systems}. All software systems are implemented in Java. While this limits, among others, the corpus of subject systems to be considered, we decided to use Java for mainly two reasons. Practically, because we can forgo code modifications for instrumentation by using off-line instrumentation (cf.~Section~\ref{sec:profiling}) for code coverage measurement. Strategically, because we can limit the influence of different measurement pipelines. We discuss this decision in more detail in the threats to validity (cf.~Section~\ref{sec:threats}). A full list of our subject systems and characteristics is presented in Table~\ref{tab:subject_systems}.

\jumper\footnote{\url{https://github.com/Sciss/jump3r/}} is a re-implementation of the LAME audio codec for MP3 in Java. In total, we have selected {\color{red}six} WAVE audio files as a workload to encode to MP3. While the selection of workloads was exploratory, we indcluded audio files with different characteristics, such as sampling rate, number of channels, and length. 

\kanzi\footnote{\url{https://github.com/flanglet/kanzi/}} is a file compression tool. For this subject system, we selected {\color{red}ten} workloads, including, among others, dedicated compression benchmarks (sets of files of different types) as well as binary and text files at different scales: a binary of the Linux kernel, a dump of its repository, XML and CSV data. 

\dconvert\footnote{\url{https://github.com/patrickfav/density-converter/}} ...

\batik ...

\jadx ...

\htwo ...

\hsqldb is a ...


\subsubsection{Configuration Sampling}
For all subject systems, we sampled a set of configurations since exhaustive assessment of the configuration space is infeasible due to combinatorial explosion. We combine several coverage-based sampling strategies and uniform random sampling into an \emph{ensemble} approach: to study the influence of single configuration options, we employ feature-wise and negative feature-wise sampling, where each feature is enabled once, or all except one, respectively. In the same way, pair-wise sampling applies this schema to study influences of two-way interactions between configuration options. Interactions of higher degree can be found accordingly, however, is computationally expensive. Instead, we augment our sample set with a random sample that is at least twice the size of the coverage-based sample subset. To achieve a uniform random sample, we used \emph{distance-based sampling}~\cite{kaltenecker_distance-based_2019}, which ensures uniform coverage of the configuration space. The variability models as well as sample sets can be found at the paper's companion web site.

\subsubsection{Coverage Profiling}\label{sec:profiling}

\subsubsection{Hardware Setup}
All experiments were conducted on compute clusters with an identical hardware setup, either Intel~Core~i5-8259U CPUs at 2.3~GHz (\jumper, \kanzi, \htwo, and \hsqldb) or Intel~Core~i7-8559U CPUs at 2.7~GHz (\dconvert, \batik, and \jadx) and 32~GB of RAM, respectively. The machines ran a headless Debian 10 installation (kernel version \mbox{4.19.0-14}). To minimize measurement noise, no additional user process were running in the background and no other than necessary packages were installed.
For all data points, we report the median across five repetitions (except for \hsqldb or \htwo), which has shown to be a good trade-off between variance and measurement effort. The reported coefficient of variation was below {\color{red}10\,\%}\change{Check number}. For the two database systems, we omitted the repetitions as in a pre-study, we found that across all benchmarks the throughput coefficient of variation (standard deviation divided by the arithmetic mean) was consistently below 5\,\%.

\subsection{Performance Spectra}
For \RQref{1.1}, we ...
{\color{teal}
\begin{itemize}
	\item Rank configurations by performance for each workload
	\item Compare Kendall's $\tau$ correlation for each pair of workloads
	\item Graphics: Heatmaps, clustered
\end{itemize}
}

\subsection{Coverage Set Similarity}
For \RQref{1.2}, we ...
\begin{itemize}
	\item For how many configurations are there workload-specific coverage differences?
	\item What is the distribution similarity (Jaccard) of lines covered per configuration across all workloads
\end{itemize}

\subsection{Correlation Analysis}
For \RQref{1.3}, we ...
\begin{itemize}
	\item Compare pairs of workloads $(w_1, w_2)$, do differences in rank of configuration $c$ correspond to 
	\begin{enumerate}
		\item any difference in coverage (similarity $< 1$)
		\item increase/decrease in coverage ($cov(w_i) \subset cov(w_j)$)

	\end{enumerate}
	\item The stronger the correlation, the more differences in code coverage could serve as a proxy for performance differences 
\end{itemize}

\clearpage
\section{Estimating Representativeness}~\label{sec:metric}

As the main target of our efforts, in the second part of our efforts, we now focus on how workload variability influences performance models. To further see, whether workload-specific coverage differences affect not only the performance of individual configurations (cf.~\RQref{1.1}), but the performance influence of each configuration option (or interaction among them), we ask the following research question:

\RQ{2.1}{Do workload-specific differences in code coverage correspond to differences in the performance influence of configuration options?}

Drawing a connection between differences in both code coverage and performance influences, depending on the workload, is a first approximation of how susceptible a performance model is to workload variability. To obtain a clearer picture of how susceptible individual configuration options and interactions among them are to workload bias, we require information about what code is related to one or more configuration option. To achieve this, we aim to infer such locations from using different workloads via differential coverage testing (cf. Section~\ref{sec:feature_location}) and ask the following research question:

\RQ{2.2}{Can we use different workloads to identify option-specific code sections via superimposition?}

To ultimately assess, whether we can estimate the bias for individual performance influences, we apply the findings from \RQref{2.2} as a baseline for feature location and correlate them with observed differences in performance influences to answer the final research question:

\RQ{2.3}{Can we use code coverage or option-code coverage as an indicator for representativeness?}

We answer the first three research questions with an empirical study in the following Section~\ref{sec:study}, followed by a further analysis in Section~\ref{sec:metric} for the remaining two research questions.


\subsection{Approximated Feature Regions}

\subsection{Representativeness Scoring}

\section{Discussion}

\section{Threats to Validity}\label{sec:threats}
\subsection{Internal Validity}\label{sec:internal_validity}
\begin{itemize}
	\item dedicated hardware to mitigate measurement noise, 
	\item five repetitions per data point, 
	\item database experiments robustness assessed in pre-study,
	\item separate experiment runs for coverage analysis and performance measurement to avoid profiling averhead (assuming determinism)
	\item load generator overhead for databases, hence omitted memory consumption and other hardware-related performance indicators 
\end{itemize}

\subsection{External Validity}\label{sec:external_validity}
\begin{itemize}
	\item Java only, but different application domains,
	\item consistent with observations from related work
\end{itemize}

\subsection{Construct Validity}\label{sec:construct_validity}
\begin{itemize}
	\item \textit{off-line instrumentatio}n vs. on-line instrumentation
	\begin{itemize}
		\item avoid compiler optimizations or overhead from instrumented source code
		\item byte code ad-hoc instrumentation could increase overhead in a separate run
		\item byte-code instrumentation with Jacoco less precise than other techniques since mapping byte code instructions back to Java source code is not always trivial (observed before~\cite{luo_2019_cova})
	\end{itemize}
	\item Coverage analysis for middle- to low-level languages face instrumentation bias
\end{itemize}

\cite{difallah_oltp_2013}

\section{Conclusion}